---
title: "San Francisco Crime Data Mining Project"
author: "Arash, Mike, Russell"
date: "May 14, 2016"
output: html_document
---
***
###Project Summary
This is the final group project for Data Mining (CST495).
We are going to use San Francisco crime data to predict how dangerous it is to be in different locations during different times.
We are going to look at the data from 911 calls. Dataset has a number of features but we are going to mostly use Police District, Time of day, Day of week and Type of crime (is it violent or not).

***

#Data Acquisition
We acquired the dataset from Kaggle.com.
We are using data from 9/7/14 to 1/10/15 which contains more than 76000 emergency calls to SFPD.
https://www.kaggle.com/c/sf-crime/data
```{r}

rawData = read.csv("/Users/wiseR3B3L/Documents/CS/R Projects/crimedata/crimetableset.csv")

```

###Loading requiered libraries and resources
```{r}

```
***
#Data Exploration
We need to explore the data to have a better understanding about the data that we are working with.
We have prepared different plot and heat maps to display how scattered is our data in different neighborhoods.

###Dataset Structure and Summary
```{r}

str(rawData)

summary(rawData)

```

***

#Data Cleaning 
###Extracting hours and Days and month from Date variable
```{r}

Hours <- format(as.POSIXct(strptime(rawData$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%H:%M")
HourOnly <- format(as.POSIXct(strptime(rawData$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%H")
MonthOnly <- format(as.POSIXct(strptime(rawData$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%m")

```

###Adding new columns to the raw data frame
```{r}

rawData["HourOnly"] = NA
rawData$HourOnly=as.numeric(HourOnly)

rawData["MonthOnly"] = NA
rawData$MonthOnly=as.numeric(MonthOnly)

```

###Adding a new column based on Category column where violent crimes get 1 and non-violent crimes get 0
```{r}

rawData["IsViolent"] = NA
rawData$IsViolent = ifelse(rawData$Category=="ASSAULT"
                       |rawData$Category=="SEX OFFENSES FORCIBLE"
                       |rawData$Category=="ROBBERY"
                       |rawData$Category=="KIDNAPPING", 1, 0)


head(rawData,5)
```

###Daily crime rate
```{r}
par(las=2,mar=c(6,3,2,2))
barplot(sort(table(rawData$DayOfWeek)), ylim = c(0, 12000), col="navy", main="Crimes By Day of Week")
```


####Number of daily assaults
```{r}
table(rawData$DayOfWeek[rawData$Category=="ASSAULT"])
par(las=2,mar=c(6,3,2,2))
barplot(sort(table(rawData$DayOfWeek[rawData$Category=="ASSAULT"])),col = "navy",ylim = c(0, 1200), main="Assaults by Day of Week")
```

###Hourly Assault Rate
```{r}
par(las=2,mar=c(3,3,2,2))
barplot(table(rawData$HourOnly[rawData$Category=="ASSAULT"]),col = "navy",ylim=c(0, 400), main="Number of Assaults Hourly", xlab="Hour of the day")
```

###Monthly Assault Rate
```{r}
par(las=2,mar=c(3,3,2,2))
barplot(table(rawData$MonthOnly[rawData$Category=="ASSAULT"]),col = "navy",ylim=c(0,700), main="Number of Assaults by Months", xlab="Month of the year")
```

###Monthly Robbery Rate
```{r}
par(las=2,mar=c(3,3,2,2))
barplot(table(rawData$MonthOnly[rawData$Category=="ROBBERY"]),col = "Navy",ylim=c(0,230), main="Robberies by the month")
```

###Grouping Assaults by Police District
```{r}
summary(rawData$PdDistrict)
table(rawData$PdDistrict[rawData$Category=="ASSAULT"])
par(las=2,mar=c(3,3,2,2))
plot(rawData$PdDistrict[rawData$Category=="ASSAULT"],ylim=c(0,1200), pch=19, main="Assaults by district")
```

