---
title: "San Francisco Crime Data Mining Project"
author: "Arash, Mike, Russell"
date: "May 14, 2016"
output: html_document
---
***
###Project Summary
This is the final group project for Data Mining (CST495).
We are going to use San Francisco crime data to predict how dangerous it is to be in different locations during different times.
We are going to look at the data from 911 calls. Dataset has a number of features but we are going to mostly use Police District, Time of day, Day of week and Type of crime (is it violent or not).

***

###Loading requiered libraries and resources
```{r}
library(rpart)
library(rpart.plot)
library(e1071)
library(rgdal)    
library(ggplot2)
library(RColorBrewer)
library(rgeos)
```


#Data Acquisition
We acquired the dataset from Kaggle.com.
We are using data from 9/7/14 to 1/10/15 which contains more than 76000 emergency calls to SFPD.
https://www.kaggle.com/c/sf-crime/data
```{r}

rawData = read.csv("/Users/wiseR3B3L/Documents/CS/R Projects/crimedata/crimetableset.csv")
SFMap = readOGR("/Users/wiseR3B3L/Downloads/Historic Police Districts/","SFPD")

```


***
#Data Exploration
We need to explore the data to have a better understanding about the data that we are working with.
We have prepared different plot and heat maps to display how scattered is our data in different neighborhoods.

###Dataset Structure and Summary
```{r}

str(rawData)

summary(rawData)

```

***

#Data Cleaning 
###Extracting hours and Days and month from Date variable
```{r}

Hours <- format(as.POSIXct(strptime(rawData$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%H:%M")
HourOnly <- format(as.POSIXct(strptime(rawData$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%H")
MonthOnly <- format(as.POSIXct(strptime(rawData$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%m")

```

###Adding new columns to the raw data frame
```{r}

rawData["HourOnly"] = NA
rawData$HourOnly=as.numeric(HourOnly)

rawData["MonthOnly"] = NA
rawData$MonthOnly=as.numeric(MonthOnly)

```

###Adding a new column based on Category column where violent crimes get 1 and non-violent crimes get 0
```{r}

rawData["IsViolent"] = NA
rawData$IsViolent = ifelse(rawData$Category=="ASSAULT"
                       |rawData$Category=="SEX OFFENSES FORCIBLE"
                       |rawData$Category=="ROBBERY"
                       |rawData$Category=="KIDNAPPING", 1, 0)

rawData$IsViolent = factor(rawData$IsViolent, labels=c("nonViolent", "violent"))
rawData$HourOnly = factor(rawData$HourOnly)

head(rawData,5)
```

###Daily crime rate
```{r}
par(las=2,mar=c(6,3,2,2))
barplot(sort(table(rawData$DayOfWeek)), ylim = c(0, 12000), col="navy", main="Crimes By Day of Week")
```


####Number of daily assaults
```{r}
table(rawData$DayOfWeek[rawData$Category=="ASSAULT"])
par(las=2,mar=c(6,3,2,2))
barplot(sort(table(rawData$DayOfWeek[rawData$Category=="ASSAULT"])),col = "navy",ylim = c(0, 1200), main="Assaults by Day of Week")
```

###Hourly Assault Rate
```{r}
par(las=2,mar=c(3,3,2,2))
barplot(table(rawData$HourOnly[rawData$Category=="ASSAULT"]),col = "navy",ylim=c(0, 400), main="Number of Assaults Hourly", xlab="Hour of the day")
```

###Monthly Assault Rate
```{r}
par(las=2,mar=c(3,3,2,2))
barplot(table(rawData$MonthOnly[rawData$Category=="ASSAULT"]),col = "navy",ylim=c(0,700), main="Number of Assaults by Months", xlab="Month of the year")
```

###Monthly Robbery Rate
```{r}
par(las=2,mar=c(3,3,2,2))
barplot(table(rawData$MonthOnly[rawData$Category=="ROBBERY"]),col = "Navy",ylim=c(0,230), main="Robberies by the month")
```

###Grouping Assaults by Police District
```{r}
summary(rawData$PdDistrict)
table(rawData$PdDistrict[rawData$Category=="ASSAULT"])
par(las=2,mar=c(7,3,2,2))
plot(rawData$PdDistrict[rawData$Category=="ASSAULT"],ylim=c(0,1200), pch=19, main="Assaults by district", col = "navy")
```

***


#HeatMap Setup
```{r}
dat = rawData
map.df <- fortify(SFMap)
centers <- SpatialPointsDataFrame(gCentroid(SFMap, byid=TRUE), SFMap@data, match.ID=FALSE)
centDF = as.data.frame(centers)
centroids <- gCentroid(SFMap, byid=TRUE)
```

##Some data cleaning for Heat Map
```{r}
Hours <- format(as.POSIXct(strptime(dat$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%H:%M")
HourOnly <- format(as.POSIXct(strptime(dat$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%H")
MonthOnly <- format(as.POSIXct(strptime(dat$Dates,"%m/%d/%Y %H:%M",tz="")) ,format = "%m")
dat["HourOnly"] = NA
dat$HourOnly=as.numeric(HourOnly)
dat["MonthOnly"] = NA
dat$MonthOnly=as.numeric(MonthOnly)
dat["IsViolent"] = NA
dat$IsViolent = ifelse(dat$Category=="ASSAULT"
                       |dat$Category=="SEX OFFENSES FORCIBLE"
                       |dat$Category=="ROBBERY"
                       |dat$Category=="KIDNAPPING", 1, 0)

```



###HeatMap of Total crime in San Francisco, May 2014 -> May 2015
```{r}
#all crime heatmap
ggplot(dat, aes(x=X, y=Y))  + geom_point(colour="red", size = 0.1)  + stat_density2d(aes(fill = ..level..), alpha=1, geom="polygon")+ scale_fill_gradientn(colours=rev(brewer.pal(11,"Spectral"))) + coord_fixed() + geom_path(data=map.df,aes(x=long, y=lat,group=group), colour="black") + geom_text(data=centDF, mapping=aes(x=x, y=y, label=district), size=3, vjust=0.5, hjust=0.5) + ggtitle("San Francisco ALL Crime Occurences Heatmap\n May 2014 - May 2015") + theme(plot.title = element_text(lineheight=2, face="bold"))
```

###HeatMap of all VIOLENT crime in San Francisco, May 2014 -> May 2015
```{r}
#violent crimes heatmap
ggplot(dat[dat$IsViolent == 1,], aes(x=X, y=Y))  + geom_point(colour="red", size = 0.1)  + stat_density2d(aes(fill = ..level..), alpha=1, geom="polygon")+ scale_fill_gradientn(colours=rev(brewer.pal(11,"Spectral"))) + coord_fixed() + geom_path(data=map.df,aes(x=long, y=lat,group=group), colour="black") + geom_text(data=centDF, mapping=aes(x=x, y=y, label=district), size=3, vjust=0.5, hjust=0.5) + ggtitle("San Francisco Violent Crime Occurences Heatmap\n May 2014 - May 2015") + theme(plot.title = element_text(lineheight=2, face="bold"))
```

###HeatMap of Violent crimes in San Francisco, from 5am to 11am  May 2014 -> May 2015
```{r}
#5am - 11am Heatmap Violent Crimes
ggplot(dat[(dat$IsViolent == 1 & dat$HourOnly > 5 & dat$HourOnly < 11),], aes(x=X, y=Y)) + geom_point(colour="red", size = 0.1)  + scale_fill_gradientn(colours=rev(brewer.pal(11,"Spectral"))) + stat_density2d(aes(fill = ..level..), alpha=1, geom="polygon") + coord_fixed() + geom_path(data=map.df,aes(x=long, y=lat,group=group), colour="black") + geom_text(data=centDF, mapping=aes(x=x, y=y, label=district), size=3, vjust=0.5, hjust=0.5) + ggtitle("San Francisco Violent Crime Occurences Heatmap\n5am -11am, May 2014 - May 2015") + theme(plot.title = element_text(lineheight=2, face="bold"))
```

###HeatMap of Violent crimes in San Francisco, from 12pm to 6pm  May 2014 -> May 2015
```{r}
#1200-1800 Heatmap Violent Crimes
ggplot(dat[(dat$IsViolent == 1 & dat$HourOnly > 12 & dat$HourOnly < 18),], aes(x=X, y=Y))  + geom_point(colour="red", size = 0.1) + stat_density2d(aes(fill = ..level..), alpha=1, geom="polygon") + scale_fill_gradientn(colours=rev(brewer.pal(11,"Spectral"))) + coord_fixed() + geom_path(data=map.df,aes(x=long, y=lat,group=group), colour="black") + geom_text(data=centDF, mapping=aes(x=x, y=y, label=district), size=3, vjust=0.5, hjust=0.5) + ggtitle("San Francisco Violent Crime Occurences Heatmap\n 12pm - 6pm, May 2014 - May 2015") + theme(plot.title = element_text(lineheight=2, face="bold"))
```

###HeatMap of Violent crimes in San Francisco, from 7pm to 5am  May 2014 -> May 2015
```{r}
ggplot(dat[(dat$IsViolent == 1 & (dat$HourOnly > 19 | dat$HourOnly < 5)),], aes(x=X, y=Y))  + geom_point(colour="red", size = 0.1) + stat_density2d(aes(fill = ..level..), alpha=1, geom="polygon") + scale_fill_gradientn(colours=rev(brewer.pal(11,"Spectral"))) + coord_fixed() + geom_path(data=map.df,aes(x=long, y=lat,group=group), colour="black") + geom_text(data=centDF, mapping=aes(x=x, y=y, label=district), size=3, vjust=0.5, hjust=0.5) + ggtitle("San Francisco Violent Crime Occurences Heatmap\n7pm - 5am, May 2014 - May 2015") + theme(plot.title = element_text(lineheight=2, face="bold"))
```

***

#Data Modeling Section

###Creating Split Function
```{r}
split_data = function(dat, frac=c(0.75, 0.25)) {
  
  k = length(frac)
  stopifnot(k > 0)
  
  n = nrow(dat)
  frac = frac/(sum(frac))
  starts = c(1, round(cumsum(frac) * n)[-k])
  ends = c(starts[-1]-1,n)
  samp = sample(1:n)
  data_sets = list()
  for (i in 1:k) {
    data_sets[[i]] = dat[samp[starts[i]:ends[i]],]
  }
  return(data_sets)
} 
```


###Spliting the date into training and test datesets
```{r}
set.seed(132)
split = split_data(rawData)
tr_dat = split[[1]]
te_dat = split[[2]]
```

***

##Naïve Bayes
Using train data to make the model and using test data to see how acurate it is
```{r}
fit = naiveBayes(IsViolent ~ ., data=tr_dat)
predicts = predict(fit, newdata=te_dat)
conf_mtx = table(predicts, te_dat$IsViolent)
```

confusion matrix and accuracy of the model
```{r}
mean(predicts == te_dat$IsViolent)
conf_mtx
```

***

##Calculating Crime Index

###Crime Index will be based on District, Day of week, hour of day and if it is violent.
```{r}
df = data.frame(table(rawData$PdDistrict, rawData$DayOfWeek, rawData$HourOnly))
quantile(df$Freq,probs=seq(0,1,length.out=20))

ViolentDF = data.frame(table(rawData$PdDistrict[rawData$IsViolent=="violent"], rawData$DayOfWeek[rawData$IsViolent=="violent"], rawData$HourOnly[rawData$IsViolent=="violent"]))
tempCrimeData = data.frame(table(rawData$PdDistrict[rawData$IsViolent=="nonViolent"], rawData$DayOfWeek[rawData$IsViolent=="nonViolent"], rawData$HourOnly[rawData$IsViolent=="nonViolent"]))

ViolentDF["WeightedFreq"]  = NA
tempCrimeData["WeightedFreq"]  = NA
tempCrimeData["ViolentWeighted"] =NA
tempCrimeData["Combined"] =NA
tempCrimeData["CrimeIndex"] =NA
df["WeightedFreq"]  = NA

ViolentDF["WeightedFreq"] = ViolentDF$Freq*.7
tempCrimeData["WeightedFreq"] = tempCrimeData$Freq*.3
tempCrimeData$ViolentWeighted = ViolentDF$WeightedFreq
tempCrimeData$ViolentWeighted <- as.numeric(tempCrimeData$ViolentWeighted )
tempCrimeData$WeightedFreq <- as.numeric(tempCrimeData$WeightedFreq)
tempCrimeData$Combined = tempCrimeData$ViolentWeighted + tempCrimeData$WeightedFreq
tempCrimeData$CrimeIndex = tempCrimeData$Combined / 62.3

crimeIndexData = data.frame(tempCrimeData$Var1, tempCrimeData$Var2, tempCrimeData$Var3, tempCrimeData$CrimeIndex)

names(crimeIndexData) <- c("PdDistrict", "DayOfWeek", "HourOnly", "CrimeIndex")
View(crimeIndexData)
```

***


##Linear Model using Crime Index
```{r}

fit = lm(CrimeIndex ~ .,data=crimeIndexData)

summary(fit)

predicted = predict(fit)
actual = crimeIndexData$CrimeIndex

rng = range(c(predicted, actual))
par(las=2,mar=c(3,3,2,2))
plot(predicted, actual, pch=20, xlim=rng, ylim=rng, xlab="predicted", ylab="actual", main = "Predicted vs Actuals Using Linear Model")
lines(c(rng[1], rng[2]), c(rng[1], rng[2]), lty=2, col="blue", lwd=1.5)

```


##Logistic Regression Model using Crime Index
```{r}

fit = glm(CrimeIndex ~ .,data=crimeIndexData, family = binomial)

summary(fit)

predicted = predict(fit, type = "response")
actual = crimeIndexData$CrimeIndex

rng = range(c(predicted, actual))
par(las=2,mar=c(3,3,2,2))
plot(predicted, actual, pch=20, xlim=rng, ylim=rng, xlab="predicted", ylab="actual", main = "Predicted vs Actuals Using Logistic Model")
lines(c(rng[1], rng[2]), c(rng[1], rng[2]), lty=2, col="blue", lwd=1.5)
```


##Regression Tree Model using Crime Index
```{r}
set.seed(132)
split = split_data(crimeIndexData)
tr_dat = split[[1]]
te_dat = split[[2]]

fit = rpart(CrimeIndex ~ ., data=tr_dat)

prp(fit, extra=1, varlen=-10, box.col="lightgreen")

predicted = predict(fit, te_dat)
errors = te_dat$CrimeIndex - predicted
rmse = sqrt(mean(errors^2))
```






